Finch: A Datastructure-Driven Array Programming Language

### 2024-03-20 main.tex
Update on Overleaf.

\ifthenelse{1=1}{
\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
}{
\newcommand{\TODO}[1]{}
\newcommand{\help}[1]{}
\newcommand{\saman}[1]{}
\newcommand{\teo}[1]{}
\newcommand{\willow}[1]{}
\newcommand{\leftoff}{}
}

@@

Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Furthermore, multi-dimensional arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler
technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
 Thus, focusing only on sparsity limits the scope needed to handle a vast array of data. Second, the arrays and loops abstraction introduced in FORTRAN supports a lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

>>>>

\ifthenelse{1=1}{
\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
\newcommand\changwan[1]{\textcolor{green}{Help Needed: #1}}
}{
\newcommand{\TODO}[1]{}
\newcommand{\help}[1]{}
\newcommand{\saman}[1]{}
\newcommand{\teo}[1]{}
\newcommand{\willow}[1]{}
\newcommand{\leftoff}{}
}

@@

Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler
technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!). So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
 Thus, focusing only on sparsity limits the scope needed to handle a vast array of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
